A Study of Common Bug Fix Patterns in Rust \thanks{Competing Interests: We have no competing interests and are funded by a Discovery Grant from Canada's Natural Science and Engineering Research Council.}

### 2023-06-06 discussion.tex
resolved todos except one

When presenting each pattern, we discussed actionability for the most important actions. Here, we discuss the implications of our most important patterns. \emph{In summary, program repair tools must understand semantics of attributes and should be performance-aware.} It may also be useful for tools to be able to automatically infer immutability.

\todo{is the most common ordering still the same after trimming?}
The two most common patterns were general patterns modifying struct fields and attributes, with 179 and 172 \todo{are those numbers stale?} instances respectively. Unfortunately, struct field modifications are neither interesting nor actionable; they are simply used to enable other changes. Attribute modifications are far more interesting, and account for the second- and fifth-most popular patterns. From a language design perspective, the fact that there are so many attribute changes marked as bug fixes illustrates the semantic importance of attributes in Rust, and \emph{implies that program repair must understand the effects of attributes}. Yet, different attributes have different, and often library-dependent, semantic meanings. Proposing specific repair techniques that reason about attributes is beyond the scope of this work, but our results suggest that they are important to develop.

The most common borrow-checker-related pattern and third-most-common pattern overall is dropping \texttt{clone()}. This pattern is performance-related; when correctly applied, it has no semantic impact on the program, but reduces resource consumption. Rust is used in situations where performance is important; the number of instances of this pattern confirms that developers do work on performance issues. This implies that \emph{program repair tools would benefit from being performance-aware}, rather than limiting themselves strictly to functional properties of the code.

The fourth- and sixth-most popular patterns, changing the type of a struct field or of a generic type parameter, are difficult to make general statements about; there are many unrelated idiosyncratic changes in these clusters. Such changes look the same, but have different purposes. One example we showed changed a \texttt{HashMap} to an \texttt{IndexMap}, preserving iteration order. That specific nondeterminism bug could be automatically repaired upon direction from the developer. Another change modified \texttt{usize} to \texttt{isize}, perhaps reflecting a requirements change to allow signed integers. IDEs could support refactorings that modify types. We find it difficult to propose further general work in this direction.

The seventh-most-popular pattern is the Rust-specific change of dropping mutability (i.e. making data immutable). Tools like ReImInfer by~\cite{HuangMDE2012} (for a Java extension), and techniques like the one proposed in~\cite{EyolfsonAbstractImmutability} for C++, aim to infer that data is immutable. In the Rust context, it is easier to make data immutable: the semantics for immutability are clear. Immutability is part of the core Rust language. Immutability escape hatches in C++ complicate issues, but in Rust, the escape hatch is the clearly-marked \texttt{unsafe} block. This pattern thus suggests that \emph{tools for inferring Rust immutability could be viable}.

We are unable to say anything meaningful about the eighth- and ninth-most popular patterns, \texttt{G.match.pattern} and \texttt{G.type.enum.variant}. They do not seem particularly helpful for automatic program repair.

@@

\todo{recap the italicized stuff from above}
\vspace*{1em}
\fbox{\parbox{0.9\textwidth}{We summarize learnings from our bug patterns as follows:
\begin{itemize}
\item one
\item two
\end{itemize}
}}

>>>>

When presenting each pattern, we discussed actionability for the most important actions. Here, we discuss the implications of our most important patterns. \emph{In summary, program repair tools must understand semantics of attributes and should be performance-aware.} It may also be useful for tools to be able to automatically infer immutability. Here, we discuss the top ten most popular patterns, 8 of which fall into the category of general patterns, and the remaining 2 are BC-related patterns.

The most common general pattern and the most popular pattern overall is modifying struct fields with 179 datapoints. Unfortunately, struct field modifications are neither interesting nor actionable; they are simply used to enable other changes.

The most common borrow-checker-related pattern and second-most-common pattern overall is dropping \texttt{clone()}. This pattern is performance-related; when correctly applied, it has no semantic impact on the program, but reduces resource consumption. Rust is used in situations where performance is important; the number of instances of this pattern confirms that developers do work on performance issues. This implies that \emph{program repair tools would benefit from being performance-aware}, rather than limiting themselves strictly to functional properties of the code.

Attribute modifications are interesting changes that account for the third- and ninth-most popular patterns. From a language design perspective, the fact that there are so many attribute changes marked as bug fixes illustrates the semantic importance of attributes in Rust, and \emph{implies that program repair must understand the effects of attributes}. Yet, different attributes have different, and often library-dependent, semantic meanings. Proposing specific repair techniques that reason about attributes is beyond the scope of this work, but our results suggest that they are important to develop.

The fourth- and fifth-most popular patterns, changing the type of a struct field or of a generic type parameter, are difficult to make general statements about; there are many unrelated idiosyncratic changes in these clusters. Such changes look the same, but have different purposes. One example we showed changed a \texttt{HashMap} to an \texttt{IndexMap}, preserving iteration order. That specific nondeterminism bug could be automatically repaired upon direction from the developer. Another change modified \texttt{usize} to \texttt{isize}, perhaps reflecting a requirements change to allow signed integers. IDEs could support refactorings that modify types. We find it difficult to propose further general work in this direction.

The sixth-most-popular pattern is the Rust-specific change of dropping mutability (i.e. making data immutable). Tools like ReImInfer by~\cite{HuangMDE2012} (for a Java extension), and techniques like the one proposed in~\cite{EyolfsonAbstractImmutability} for C++, aim to infer that data is immutable. In the Rust context, it is easier to make data immutable: the semantics for immutability are clear. Immutability is part of the core Rust language. Immutability escape hatches in C++ complicate issues, but in Rust, the escape hatch is the clearly-marked \texttt{unsafe} block. This pattern thus suggests that \emph{tools for inferring Rust immutability could be viable}.

We are unable to say anything meaningful about the seventh- and eighth-most popular patterns, \texttt{G.match.pattern} and \texttt{G.type.enum.variant}. They do not seem particularly helpful for automatic program repair.

@@

\todo{recap the italicized stuff from above}
\vspace*{1em}
\fbox{\parbox{0.9\textwidth}{We summarize learnings from our bug patterns as follows:
\begin{itemize}
\item Being performance-awareness would be advantageous for program repair tools.
\item Understanding the effects of attributes can guide program repair.
\item Tools for inferring Rust immutability could be viable.
\end{itemize}
}}

### 2023-06-06 methodology.tex
reread for usefulness
resolved todos except one

In this section, we describe our methodology, providing full details in later subsections. Ruxanne uses a pipeline for finding classes of pervasive patterns in Rust programs. This pipeline parses program versions, computes differences between them, embeds them in fixed-size datapoints, and then clusters these datapoints. After conducting a manual analysis on the obtained clusters, we refine them to obtain our proposed bug fix patterns.

@@

\subsubsection{\label{sec:manual_analysis_cluster_selection}Manual Analysis for cluster selection}
% 3 types, refactoring dropped

Like \cite{yang2022mining}, and following the definitions of \cite{cotroneo2019analyzing}, we divide the clusters in three different groups:

@@

For each cluster, we read the code and the bug reports of the datapoints (if they existed). If we judged that the members of a cluster fit into a pattern, we wrote a natural language description of that cluster. Considering the descriptions, we linked similar clusters together as possible candidates for merging. We considered these candidates and carried out appropriate ones. Finally, we selected 20 of the remaining clusters---the ones that manifested more-specific changes. Sections~\ref{sec:common_patterns}~and~\ref{sec:bc_patterns} present our clusters.

>>>>

In this section, we provide an overview of our methodology, providing full details in later subsections. Ruxanne uses a pipeline for finding classes of pervasive patterns in Rust programs. This pipeline parses program versions, computes differences between them, embeds them in fixed-size datapoints, and then clusters these datapoints. After conducting a manual analysis on the obtained clusters, we refine them to obtain our proposed bug fix patterns.

@@

\subsubsection{\label{sec:manual_analysis_cluster_selection}Manual Analysis for cluster selection}
% 3 types, refactoring dropped
Like \cite{yang2022mining}, and following the definitions of \cite{cotroneo2019analyzing}, we divide the clusters in three different groups:

@@

For each cluster, we read the code and the bug reports of the datapoints (if they existed). If we judged that the members of a cluster fit into a pattern, we wrote a natural language description of that cluster. Considering the descriptions, we linked similar clusters together as possible candidates for merging, and removed the datapoints that did not fit into the first two groups listed above. We considered these candidates and carried out appropriate ones. Finally, we selected 20 of the remaining clusters---the ones that manifested more-specific changes. Sections~\ref{sec:common_patterns}~and~\ref{sec:bc_patterns} present our clusters.

### 2023-06-06 results.tex
resolved todos except one

\noindent\textbf{Description:} This is a performance-related \todo{is it only perf?} change that in all instances (in our benchmarks) has been proposed by Clippy---commit messages indicated the use of Clippy for this fix. The change applies on an object prior to calling \verb+to_string()+ on that object. In this change, a dereference operator is added to that object. The reason behind the change is that the object is a reference type of \verb+T+, while \verb+T+ directly implements \verb+to_string()+. Adding the dereference makes the compiler use the specialized implementation of \verb+to_string()+ and not go through slower string formatting methods\footnote{\url{https://rust-lang.github.io/rust-clippy/master/index.html\#inefficient\_to\_string}}.

>>>>

\noindent\textbf{Description:} This is a change that in all instances (in our benchmarks) has been proposed by Clippy---commit messages indicated the use of Clippy for this fix. The change applies on an object prior to calling \verb+to_string()+ on that object. In this change, a dereference operator is added to that object. The reason behind the change is that the object is a reference type of \verb+T+, while \verb+T+ directly implements \verb+to_string()+. Adding the dereference makes the compiler use the specialized implementation of \verb+to_string()+ and not go through slower string formatting methods\footnote{\url{https://rust-lang.github.io/rust-clippy/master/index.html\#inefficient\_to\_string}}. This change is aimed at improving performance, as the functionality remains unchanged.
